---
title: "Mini Project 3: CLASSIFICATION OF MACLEISH TRAILS BY DIFFERENT FACTORS"
author: "Cathy Lee, Ha Cao, Lizette Carpenter"
date: "7 April 2018"
output: 
  html_document:
    theme: united
    code_folding: hide
---

### Introduction and setup

Please see our Github repository for more details on this project.[^github]

Smith maintains Macleish Field Station in West Whately, MA that provides opportunities for faculty and students to pursue environmental research, outdoor education, and low-impact creation. The data we use in this project is from `macleish` package retrieved from the station, and you can read official documentation about it [here](https://cran.r-project.org/web/packages/macleish/macleish.pdf).

**Problem Statement**: After talking to Professor Bertone-Johnson about what kinds of classification of the trails in use by field station visitors he would be interested to know, in addition to their **difficulty level**, we have decided to classify the existing trails based on their **biodiversity** (forests that they pass through) and **the number of Arts Afield Sites** within their viewing distance as well. 
```{r message = FALSE, warning = FALSE}
# load the libraries and data
#devtools::install_github("beanumber/macleish")
library(tidyverse)
library(sf)
library(macleish)
library(leaflet)
proj4_aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
```

### Classification of existing trails by level of difficulty based on length and elevation change

According to this [rating system](https://www.nps.gov/shen/planyourvisit/how-to-determine-hiking-difficulty.htm) by National Park Service, all these trails would be rated as "Easy", so we developed our own heuristic that makes the difficulty level relative to Macleish only:

* Length: We rate this in the scale of 10. The longest trail (2575.217 m) has length score 10/10. We calculate the scores for the other trails by dividing their length by the biggest length and multiplying by 10, and round up to integers.

* Elevation: This is broken down into 2 factors: the number of times a trail changes its elevation and the mean change in height each time (by dividing the difference between the maximum and minimum elevation by the number of changes in elevation). Each of these 2 factors are measured in the scale of 5, so they sum up to the scale of 10. The more often a trail changes its elevation, and/or the bigger the changes are, the more difficult it is for hikers/bikers. 

Then we sum the partial scores into the total score in the scale of 20:

* From 0 - 7: Easy Trail
* From 8 - 14: Moderate Trail
* From 15- 20: Hard Trail
```{r message = FALSE, warning = FALSE}
# create the length variable of the trails
trails <- macleish_layers[["trails"]] %>%
  mutate(computed_length = st_length(geometry))

# group trails with the same name
trails_combined <- trails %>%
  group_by(name) %>%
  summarize(color = first(color),
            length = sum(computed_length))

# join the two sf data frames to find intersection between trails and contours
# ELEV_M is elevation above or below sea level, in meters
trails_intersect <- st_intersection(trails_combined,
                macleish_layers[["contours_3m"]]) %>%
  st_cast("POINT") 

# summarize to see for each trail, how many times it changes its elevation, what is the highest/lowest contour, and what is the mean change in height
trails_change_elevation <- trails_intersect %>%
  group_by(name) %>%
  summarize(num_change = n(),
            min_height = min(ELEV_M),
            max_height = max(ELEV_M)) %>%
  mutate(mean_change = (max_height - min_height)/num_change)

# join trails_change_Elevation back to trails_intersect
trails_intersect <- st_join(trails_intersect, trails_change_elevation)

# transform the length variable from factor to numeric 
trails_intersect$length <- as.numeric(trails_intersect$length, "")

# calculate the scores
max_length <- max(trails_intersect$length)
max_num_change <- max(trails_intersect$num_change)
max_mean_change <- max(trails_intersect$mean_change)

trails_intersect <- trails_intersect %>%
  mutate(score_length = round(length*10/max_length),
         score_num = round(num_change*5/max_num_change),
         score_mean = round(mean_change*5/max_mean_change)) %>%
  mutate(total_score = score_length + score_num + score_mean) %>%
  mutate(level = if_else(total_score > 14, "hard", if_else(total_score > 7, "moderate", "easy"))) 

# summarize into a shorter version
trails_short <- trails_intersect %>%
  group_by(name.x) %>%
  summarize(score = first(total_score),
            level = first(level))

st_geometry(trails_short) <- NULL

# join trails_short back to trails_combined
trails_mapped <- full_join(trails_combined, trails_short, by = c("name" = "name.x")) 

# change orders of factors in the variable `level`
trails_mapped$level = factor(trails_mapped$level, levels = c("easy", "moderate", "hard"))

# plot the trails and their levels into the map
pal <- colorFactor(palette = c("darkorange", "blue", "red"), domain = trails_mapped$level)

pal_contours <- colorNumeric(palette = "viridis", domain = macleish_layers[["contours_3m"]]$ELEV, reverse = TRUE)

leaflet(data = trails_mapped, width = "100%") %>%
  addTiles() %>%
  addPolylines(data = macleish_layers[["contours_3m"]], color = ~pal_contours(ELEV_M), weight = 2, group = "10' Contours") %>%
  addLegend("bottomright", pal = pal_contours, values = ~macleish_layers[["contours_3m"]]$ELEV_M, title = "10' Contours Elevation in meters", opacity = 1) %>%
  addPolylines(color = ~pal(level), label = ~name, group = "Trails", opacity = 0.8) %>%
  addLegend("bottomright", pal = pal, values = ~level, title = "Difficulty Level", opacity = 1) %>%
  addLayersControl(
    overlayGroups = c("Trails", "10' Contours"),
    options = layersControlOptions(collapsed = FALSE)
  ) 
```
From the map, the Easy Out trail is weird because its two segments are not continuous. After we asked Professor Bertone-Johnson, it turns out that they're just little connectors that allow people to get out to the Eastern Loop trail without continuing around the loops of the other trails.

We can see there is a positive correlation between the lengths of the trails and the difficulty level. Obviously, length contribute half of the difficulty in our heuristic, but also the longer trails tend to be closer to the inside of Macleish, which mean they are more likely to cross higher contours. Also, because they are longer, they cut a lot more different contours constantly, which means they change their elevations more often, making it harder to hike/bike on those trails. 

### Classification of trails by biodiversity/vegetation

Our heuristic to rate the biodiversity/vegetation of trails:

* Number of forests each trail comes across: We rate this in the scale of 10, and conduct rating in the same way as for difficulty level. This indicates how many times the trail changes its natural scenery.

* Number of distinct forest types each trail passes through: We rate this in the scale of 10. This shows how diverse the vegetation along the trail is. 

* Average distance between forests: We rate this in the scale of 4, because even the longest average distance 367.89 m (length/number of forests) is pretty small and within walking distance, so we weigh this factor less. The trail with the shortest average distance has the score of 4. We want to include this because we want to make the number of forests along the trail relative to its length, because obviously longer trails are likely to go through more forests. 

Then we sum the partial scores into the total score in the scale of 20:

* From 0 - 8: Least Biologically Diverse Trail
* From 9 - 16: More Biologically Diverse Trail
* From 17- 24: Most Biologically Diverse Trail
```{r message = FALSE, warning = FALSE, echo, include = FALSE}
# join the two sf data frames to find intersection between trails and forests
trails_forests <- st_intersection(trails_combined,
                macleish_layers[["forests"]]) 
addNA(trails_forests$Sheet1__Na)
trails_forests <- trails_forests %>%
  rename(forest_type = Sheet1__Na) # rename to make the variable name more telling

# find how many forests each trail passes by and how many unique forest types it passes by
trails_forests$length <- as.numeric(trails_forests$length, "")
trails_forests <- trails_forests %>%
  group_by(name) %>%
  summarize(distinct_veg_type = n_distinct(VegType_21),
            distinct_forest_type = n_distinct(forest_type),
            num_forests_crossed = n(), 
            length = first(length)) %>%
# add a variable to show distance hikers/bikers have to go to see a forest
  mutate(distance_to_forest = length/num_forests_crossed)

# calculate the scores
max_num_forests <- max(trails_forests$num_forests_crossed)
max_num_distinct_forests <- max(trails_forests$distinct_forest_type)
min_avg_distance <- min(trails_forests$distance_to_forest)

trails_forests <- trails_forests %>%
  mutate(score_num = round(num_forests_crossed*10/max_num_forests),
         score_distinct = round(distinct_forest_type*10/max_num_distinct_forests),
         score_avg_distance = round(min_avg_distance*4/distance_to_forest)) %>%
  mutate(total_score = score_num + score_distinct + score_avg_distance) %>%
  mutate(level = if_else(total_score > 16, "Most Biologically Diverse Trail", if_else(total_score > 8, "Moderately Biologically Diverse Trail", "Least Biologically Diverse Trail"))) %>%
  mutate(num_forests_range = if_else(num_forests_crossed > 5, "6 - 10 forests", "1 - 5 forests")) %>%
  mutate(num_distinct_forests_range = if_else(distinct_forest_type > 4, "5 - 7 types of forests", "1 - 4 types of forests"))

st_geometry(trails_forests) <- NULL

# join trails_forests back to trails_combined
trails_forests <- full_join(trails_combined, trails_forests, by = c("name" = "name")) 
# change orders of factors in the variable `level`, 'num_forests_range`, and `num_distinct_forests_range`
trails_forests$level = factor(trails_forests$level, levels = c("Least Biologically Diverse Trail", "Moderately Biologically Diverse Trail", "Most Biologically Diverse Trail"))
trails_forests$num_forests_range = factor(trails_forests$num_forests_range, levels = c("1 - 5 forests", "6 - 10 forests"))
trails_forests$num_distinct_forests_range = factor(trails_forests$num_distinct_forests_range, levels = c("1 - 4 types of forests", "5 - 7 types of forests"))
```

```{r}
# plot the trails and their levels into the map
pal_level <- colorFactor(palette = c("red", "blue", "darkgreen"), domain = trails_forests$level)
pal_range <- colorFactor(palette = c("red", "blue"), domain = trails_forests$num_forests_range)
pal_distinct <- colorFactor(palette = c("red", "blue"), domain = trails_forests$num_distinct_forests_range)

leaflet(data = trails_forests, width = "100%") %>%
  addTiles() %>%
  addPolygons(data = macleish_layers[["forests"]], color = "lightgreen", weight = 1, fillOpacity = 0.5, group = "Forests", label = ~Sheet1__Na) %>%
  addPolylines(color = ~pal_level(level), label = ~name, group = "Overall Biodiversity Level", opacity = 0.8) %>%
  addLegend("bottomright", pal = pal_level, values = ~level, title = "Biodiversity Level", opacity = 1) %>%
  addPolylines(color = ~pal_range(num_forests_range), label = ~name, group = "Number of Forests", opacity = 0.8) %>%
  addLegend("bottomright", pal = pal_range, values = ~num_forests_range, title = "Number of Forests crossed", opacity = 1) %>%
  addPolylines(color = ~pal_distinct(num_distinct_forests_range), label = ~name, group = "Number of Distinct Forest Types", opacity = 0.8) %>%
  addLegend("bottomright", pal = pal_distinct, values = ~num_distinct_forests_range, title = "Number of Distinct Forest Types seen", opacity = 1) %>%
  addLayersControl(
    overlayGroups = c("Overall Biodiversity Level", "Number of Forests", "Number of Distinct Forest Types", "Forests"),
    options = layersControlOptions(collapsed = FALSE)
  ) 
```

From the map, we can see that the most biologically diverse trails are the looping ones like Eastern/Western loops. The least biologically diverse trails are part of Easy Out trail, which makes perfect sense because they are so short and are just connectors, meaning they were not built for the purpose of exploring the nature, but to help facilitate the traffic. 

### Classification of trails by the number of Arts Afield Sites within viewing distance

For this, we used the external data about Arts Afield Sites that were provided by Professor Bertone-Johnson. We'll put these data into our Github so that you can run the codes later. 

The heuristic for this classificiation is simpler because the number of Arts Afield Sites within viewing distance (50 meters) for each trail is pretty small, so we just categorize the trails into 0, 1-2, and 3-4 sites. If visitors take the trails that are close to a lot of arts sites, it will be more convenient for them to visit these sites if they are interested in arts.  
```{r message = FALSE, warning = FALSE, include = FALSE}
# read in the Arts Afield Sites shapefile
arts_afield <- st_read("Arts_Afield/Arts_Afield_20160406.shp")
afield_es <- st_transform(arts_afield, crs = st_crs(trails_mapped))

# create buffers of 300m around each Arts Afield site
afield_buffer <- afield_es %>%
  st_transform(proj4_aea) %>%
  st_buffer(dist = 50) %>%
  st_transform(4326)

# calculate intersection of field with Arts Afield Site buffer
trail_afield_intersect <- st_intersection(trails, afield_buffer) 

# calculate the number of intersections each trail has with Arts Afield Sites
# In other words, how many Arts Afield Sites are within viewing distance of each trail?
trail_afield_num_intersect <- trail_afield_intersect %>%
  group_by(name) %>% 
  summarise(num_intersections = n())

st_geometry(trail_afield_num_intersect) <- NULL

# join trail_afield_plot with trails so we aren't just plotting trails within the buffer area
trail_afield_full <- full_join(trails_combined, trail_afield_num_intersect, by = c("name" = "name"))

# replace null values with 0
trail_afield_mapped <- trail_afield_full %>% 
  select(-color) 
trail_afield_mapped[is.na(trail_afield_mapped)] <- 0

# categorise the number of intersections
trail_afield_mapped <- trail_afield_mapped %>%
  mutate(level = if_else(num_intersections >= 3, "3-4 Arts Afield Sites", if_else(num_intersections >= 1, "1-2 Arts Afield Sites", "0 Arts Afield Sites")))

# change the order of factors in the variable `level`
trail_afield_mapped$level = factor(trail_afield_mapped$level, levels=c("0 Arts Afield Sites", "1-2 Arts Afield Sites", "3-4 Arts Afield Sites"))
# plot the trails and their levels into the map
pal <- colorFactor(palette = c("darkorange", "blue", "red"), domain = trail_afield_mapped$level)
```

```{r}
# plot the data
leaflet(data = trail_afield_mapped, width = "100%") %>%
  addTiles() %>%
  addPolylines(color = ~pal(level), label = ~name, group="Trails", opacity=1.0) %>%
  addLegend("bottomright", pal = pal, values = ~level) %>%
  addMarkers(data=afield_es, popup = ~ident, group="Arts Afield Sites") %>%
  addPolygons(data = afield_buffer, group="Arts Afield Sites") %>%
  addLayersControl(
    overlayGroups = c("Trails", "Arts Afield Sites"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Recommendations

Macleish is building a new kiosk for the field station and will be posting materials that are useful to visitors. **Our classifications by difficulty level, biodiversity, and arts sites can be used to build a detailed, all-inclusive guide for visitors**. They might be out for a bird watching walk and be interested in biodiversity of the habitats they will be seeing. They might be artists interested in the Arts Afield sites. They might be up for a challenge in their hike or out for a leisurely stroll. A personalized and detailed tour guide with these classifications will help visitors make more informed choices, which can optimize the use of their time and provide them with a more individual and meaningful experience. Our classifications are highly scalable, meaning that you can do similar classifications for other factors once more data about Macleish are available. To implement this guidebook for visitors will not cost much resource such as time and money, but can create great impact by making it much easier and more enjoyable for people, including researchers, students, etc. to visit Macleish, thereby attracting more visitors to the place. For a field station like Macleish, being able to attract a lot of visitors is key to its sustainability, development, and contribution to academia. 

[^github]: https://github.com/cathyleee/sds192-mp3
