---
title: "mp3"
author: "Cathy Lee, Lizette Carpenter, Ha Cao"
date: "7 April 2018"
output: html_document
---

```{r message=FALSE}
# load the libraries and data 
library(tidyverse)
library(sf)
library(macleish)
# packageVersion("macleish")
```

```{r}
# explore the data
names(macleish_layers)
# trails is existing trails shapefile
# forests is vegetation type shapefile
# streams and wetlands are hydrology shapefile
# challenge_courses is challenge course element locations shapefile
# boundary is property boundary shapefile
```

```{r}
contour_elevation <- macleish_layers[["contours_3m"]] %>%
  st_transform(4326)
```

```{r}
trails <- macleish_layers[["trails"]] %>%
  mutate(computed_length = st_length(geometry))
```

```{r}
# group trails with the same name
trails_combined <- trails %>%
  group_by(name) %>%
  summarize(color = first(color),
            length = sum(computed_length),
            geometry = st_union(geometry))
```

```{r}
# try out joining the two data frames
trails_elevation <- st_intersection(macleish_layers[["trails"]], contour_elevation) %>%
  group_by(name)
```

This is how we are going to classify the difficulty of the hiking trails

- Easy:


how do we link contour_elevation and trails? should we join them using st_intersection or st_join or what else???
trails have duplicate names with different numbers -- how should we interpret them?
should we use data other than the two suggested supporting data
what does the color in trails mean?
national park service has rated these lengths as easy, should we create our own scale? 
how can we know if there is a bridge? should we consider just length and elevation or include other things???
