---
title: "Mini Project 3: CLASSIFICATION OF MACLEISH TRAILS BY DIFFERENT FACTORS"
author: "Cathy Lee, Ha Cao, Lizette Carpenter"
date: "7 April 2018"
output: 
  html_document:
    theme: united
    code_folding: hide
---

### Introduction

Please see our Github repository[^github] for more details on this project.

**Problem Statement:** Besides assigning difficulty levels to MacLeish trails based on length and elevation, we decided to classify the trails based on biodiversity and proximity to Arts Afield Sites after talking to Professor Bertone-Johnson about other possible classifications.
```{r message=FALSE, warning=FALSE}
# load the libraries and data
library(tidyverse)
library(sf)
library(macleish)
library(leaflet)
proj4_aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
```

### Classification of trails by difficulty level

We combined data from `trails` and `contours_3m` layers in the `macleish` package. According to this [rating system](https://www.nps.gov/shen/planyourvisit/how-to-determine-hiking-difficulty.htm) by National Park Service, all MacLeish trails would be rated "Easy", so we developed our own heuristic for the MacLeish trails. The score for each trail is calculated relative to the highest score, as explained in the Length category:

* **Length:** Rated out of 10; the longest trail (2575.217m) has a score of 10, and the scores of other trails are calculated by dividing their length by the largest length, multiplying by 10, and rounding up to an integer. 
* **Elevation:** Rated out of 10.
    - Average change in elevation: Rated out of 5; the trail with the biggest average change in elevation scores 5.
    - Number of elevation changes: Rated out of 5; the trail with the most elevation changes scores 5.

We summed the partial scores and scored difficulty out of 20:

* 0-7: Easy 
* 8-14: Moderate 
* 15-20: Hard 

```{r message = FALSE, warning = FALSE}
# create the length variable of the trails
trails <- macleish_layers[["trails"]] %>%
  mutate(computed_length = st_length(geometry))

# group trails with the same name
trails_combined <- trails %>%
  group_by(name) %>%
  summarize(color = first(color),
            length = sum(computed_length))

# join the two sf data frames to find intersection between trails and contours
# ELEV_M is elevation above or below sea level, in meters
trails_intersect <- st_intersection(trails_combined,
                macleish_layers[["contours_3m"]]) %>%
  st_cast("POINT") 

# summarize to see for each trail, how many times it changes its elevation, what is the highest/lowest contour, and what is the mean change in height
trails_change_elevation <- trails_intersect %>%
  group_by(name) %>%
  summarize(num_change = n(),
            min_height = min(ELEV_M),
            max_height = max(ELEV_M)) %>%
  mutate(mean_change = (max_height - min_height)/num_change)

# join trails_change_Elevation back to trails_intersect
trails_intersect <- st_join(trails_intersect, trails_change_elevation)

# transform the length variable from factor to numeric 
trails_intersect$length <- as.numeric(trails_intersect$length, "")

# calculate the scores
max_length <- max(trails_intersect$length)
max_num_change <- max(trails_intersect$num_change)
max_mean_change <- max(trails_intersect$mean_change)

trails_intersect <- trails_intersect %>%
  mutate(score_length = round(length*10/max_length),
         score_num = round(num_change*5/max_num_change),
         score_mean = round(mean_change*5/max_mean_change)) %>%
  mutate(total_score = score_length + score_num + score_mean) %>%
  mutate(level = if_else(total_score > 14, "hard", if_else(total_score > 7, "moderate", "easy"))) 

# summarize into a shorter version
trails_short <- trails_intersect %>%
  group_by(name.x) %>%
  summarize(score = first(total_score),
            level = first(level))

st_geometry(trails_short) <- NULL

# join trails_short back to trails_combined
trails_mapped <- full_join(trails_combined, trails_short, by = c("name" = "name.x")) 

# change orders of factors in the variable `level`
trails_mapped$level = factor(trails_mapped$level, levels = c("easy", "moderate", "hard"))

# plot the trails and their levels into the map
pal <- colorFactor(palette = c("darkorange", "blue", "red"), domain = trails_mapped$level)

pal_contours <- colorNumeric(palette = "viridis", domain = macleish_layers[["contours_3m"]]$ELEV, reverse = TRUE)

leaflet(data = trails_mapped, width = "100%") %>%
  addTiles() %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topography") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>%
  addProviderTiles("Stamen.TonerLite", group = "Toner Lite") %>%
  addPolylines(data = macleish_layers[["contours_3m"]], color = ~pal_contours(ELEV_M), weight = 2, group = "10' Contours") %>%
  addLegend("bottomleft", pal = pal_contours, values = ~macleish_layers[["contours_3m"]]$ELEV_M, title = "10' Contours Elevation in meters", opacity = 1) %>%
  addPolylines(color = ~pal(level), label = ~name, group = "Trails", opacity = 0.8) %>%
  addLegend("bottomright", pal = pal, values = ~level, title = "Difficulty Level", opacity = 1) %>%
  addLayersControl(
    baseGroups = c("OpenStreetMap", "Topography", "Satellite", "Toner Lite"),
    overlayGroups = c("Trails", "10' Contours"),
    options = layersControlOptions(collapsed = FALSE)
  ) 
```
The map shows an anomaly: two segments in the Easy Out trail are not continuous; after consulting Professor Bertone-Johnson, we found out that they are connectors allowing easy exit from looping trails.

The longer trails tend to be closer to the inside of MacLeish, and thus tend to cross higher contours. Due to their length, they cut across more contours, changing elevation more often, making it harder to hike/bike on those trails.

### Classification of trails by biodiversity

For this category, we used data from the `forests` layer. The scores were calculated relative to the highest score in each category, similar to our calculations for difficulty level. Our heuristic to rate the biodiversity of trails is:

* **Number of forests:** Rated out of 10; the trail crossing the most forests scores 10. How many times the trails experience a scenery change.
* **Number of distinct vegetation types:** Rated out of 10; the trail with the most biodiversity scores 10. How diverse the vegetation along the trail is.
* **Average distance between forests:** Rated out of 4; the trail with the shortest average distance between forests scores 4. Even the longest average distance, 367.89m, is small and within walking distance. Hence, we gave this factor less weight. We included this because we wanted to make the number of forests along the trail relative to its length, because longer trails are likely to cross more forests.

We summed the partial scores and scored biodiversity out of 20:

* 0-8: Least Biologically Diverse 
* 9-16: More Biologically Diverse 
* 17-24: Most Biologically Diverse 
```{r message = FALSE, warning = FALSE, echo, include = FALSE}
# join the two sf data frames to find intersection between trails and forests
trails_forests <- st_intersection(trails_combined,
                macleish_layers[["forests"]]) 
addNA(trails_forests$Sheet1__Na)
trails_forests <- trails_forests %>%
  rename(forest_type = Sheet1__Na) # rename to make the variable name more telling

# find how many forests each trail passes by and how many unique forest types it passes by
trails_forests$length <- as.numeric(trails_forests$length, "")
trails_forests <- trails_forests %>%
  group_by(name) %>%
  summarize(distinct_veg_type = n_distinct(VegType_21),
            distinct_forest_type = n_distinct(forest_type),
            num_forests_crossed = n(), 
            length = first(length)) %>%
# add a variable to show distance hikers/bikers have to go to see a forest
  mutate(distance_to_forest = length/num_forests_crossed)

# calculate the scores
max_num_forests <- max(trails_forests$num_forests_crossed)
max_num_distinct_forests <- max(trails_forests$distinct_forest_type)
min_avg_distance <- min(trails_forests$distance_to_forest)

trails_forests <- trails_forests %>%
  mutate(score_num = round(num_forests_crossed*10/max_num_forests),
         score_distinct = round(distinct_forest_type*10/max_num_distinct_forests),
         score_avg_distance = round(min_avg_distance*4/distance_to_forest)) %>%
  mutate(total_score = score_num + score_distinct + score_avg_distance) %>%
  mutate(level = if_else(total_score > 16, "Most Biologically Diverse Trail", if_else(total_score > 8, "Moderately Biologically Diverse Trail", "Least Biologically Diverse Trail"))) %>%
  mutate(num_forests_range = if_else(num_forests_crossed > 5, "6 - 10 forests", "1 - 5 forests")) %>%
  mutate(num_distinct_forests_range = if_else(distinct_forest_type > 4, "5 - 7 types of forests", "1 - 4 types of forests"))

st_geometry(trails_forests) <- NULL

# join trails_forests back to trails_combined
trails_forests <- full_join(trails_combined, trails_forests, by = c("name" = "name")) 
# change orders of factors in the variable `level`, 'num_forests_range`, and `num_distinct_forests_range`
trails_forests$level = factor(trails_forests$level, levels = c("Least Biologically Diverse Trail", "Moderately Biologically Diverse Trail", "Most Biologically Diverse Trail"))
trails_forests$num_forests_range = factor(trails_forests$num_forests_range, levels = c("1 - 5 forests", "6 - 10 forests"))
trails_forests$num_distinct_forests_range = factor(trails_forests$num_distinct_forests_range, levels = c("1 - 4 types of forests", "5 - 7 types of forests"))
```

```{r}
# plot the trails and their levels into the map
pal_level <- colorFactor(palette = c("red", "blue", "darkgreen"), domain = trails_forests$level)
pal_range <- colorFactor(palette = c("red", "blue"), domain = trails_forests$num_forests_range)
pal_distinct <- colorFactor(palette = c("red", "blue"), domain = trails_forests$num_distinct_forests_range)

leaflet(data = trails_forests, width = "100%") %>%
  addTiles() %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topography") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>%
  addProviderTiles("Stamen.TonerLite", group = "Toner Lite") %>%
  addPolygons(data = macleish_layers[["forests"]], color = "lightgreen", weight = 1, fillOpacity = 0.5, group = "Forests", label = ~Sheet1__Na) %>%
  addPolylines(color = ~pal_level(level), label = ~name, group = "Overall Biodiversity Level", opacity = 0.8) %>%
  addLegend("bottomleft", pal = pal_level, values = ~level, title = "Biodiversity Level", opacity = 1) %>%
  addPolylines(color = ~pal_range(num_forests_range), label = ~name, group = "Number of Forests", opacity = 0.8) %>%
  addLegend("bottomleft", pal = pal_range, values = ~num_forests_range, title = "Number of Forests crossed", opacity = 1) %>%
  addPolylines(color = ~pal_distinct(num_distinct_forests_range), label = ~name, group = "Number of Distinct Forest Types", opacity = 0.8) %>%
  addLegend("bottomleft", pal = pal_distinct, values = ~num_distinct_forests_range, title = "Number of Distinct Forest Types seen", opacity = 1) %>%
  addLayersControl(
    baseGroups = c("OpenStreetMap", "Topography", "Satellite", "Toner Lite"),
    overlayGroups = c("Overall Biodiversity Level", "Number of Forests", "Number of Distinct Forest Types", "Forests"),
    options = layersControlOptions(collapsed = FALSE)
  ) 
```
The most biologically diverse trails are the looping ones, such as the Eastern/Western loops. The least biologically diverse trails are segments of the Easy Out trail, which makes sense because they are short connectors, and therefore were not built for the purpose of exploring nature, but to help facilitate traffic.

### Classification of trails by the number of Arts Afield Sites within viewing distance

Arts Afield Sites are locations for artistic work and reflection at MacLeish, and are useful for humanities-inclined visitors. We received this data from Professor Bertone-Johnson. The most art sites that were in viewing distance from a trail was 4. Thus, this category was rated out of 4.

* 0 Arts Afield Sites
* 1-2 Arts Afield Sites
* 3-4 Arts Afield Sites
```{r message = FALSE, warning = FALSE, include = FALSE}
# read in the Arts Afield Sites shapefile
arts_afield <- st_read("Arts_Afield/Arts_Afield_20160406.shp")
afield_es <- st_transform(arts_afield, crs = st_crs(trails_mapped))

# create buffers of 300m around each Arts Afield site
afield_buffer <- afield_es %>%
  st_transform(proj4_aea) %>%
  st_buffer(dist = 50) %>%
  st_transform(4326)

# calculate intersection of field with Arts Afield Site buffer
trail_afield_intersect <- st_intersection(trails, afield_buffer) 

# calculate the number of intersections each trail has with Arts Afield Sites
# In other words, how many Arts Afield Sites are within viewing distance of each trail?
trail_afield_num_intersect <- trail_afield_intersect %>%
  group_by(name) %>% 
  summarise(num_intersections = n())

st_geometry(trail_afield_num_intersect) <- NULL

# join trail_afield_plot with trails so we aren't just plotting trails within the buffer area
trail_afield_full <- full_join(trails_combined, trail_afield_num_intersect, by = c("name" = "name"))

# replace null values with 0
trail_afield_mapped <- trail_afield_full %>% 
  select(-color) 
trail_afield_mapped[is.na(trail_afield_mapped)] <- 0

# categorise the number of intersections
trail_afield_mapped <- trail_afield_mapped %>%
  mutate(level = if_else(num_intersections >= 3, "3-4 Arts Afield Sites", if_else(num_intersections >= 1, "1-2 Arts Afield Sites", "0 Arts Afield Sites")))

# change the order of factors in the variable `level`
trail_afield_mapped$level = factor(trail_afield_mapped$level, levels=c("0 Arts Afield Sites", "1-2 Arts Afield Sites", "3-4 Arts Afield Sites"))
# plot the trails and their levels into the map
pal <- colorFactor(palette = c("darkorange", "blue", "red"), domain = trail_afield_mapped$level)
```

```{r}
# plot the data
leaflet(data = trail_afield_mapped, width = "100%") %>%
  addTiles() %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topography") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>%
  addProviderTiles("Stamen.TonerLite", group = "Toner Lite") %>%
  addPolylines(color = ~pal(level), label = ~name, group="Trails", opacity=1.0) %>%
  addLegend("bottomright", pal = pal, values = ~level, title = "Number of Arts Afield Sites Within 50 meters", opacity = 1) %>%
  addMarkers(data=afield_es, popup = ~ident, group="Arts Afield Sites") %>%
  addPolygons(data = afield_buffer, group="Arts Afield Sites") %>%
  addLayersControl(
    baseGroups = c("OpenStreetMap", "Topography", "Satellite", "Toner Lite"),
    overlayGroups = c("Trails", "Arts Afield Sites"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Actionable Recommendations

MacLeish is building a kiosk for the field station, and will be posting materials that are useful to visitors. Our classifications can be used to build a detailed guidebook for visitors, including Smithies looking to further their environmental learning. Visitors can select trails with a suitable difficulty level. Organisations looking for opportunities to bond can choose “Easy” trails, while those looking to hone their fitness can choose “Hard” trails. Professors teaching plant biology would be interested in the biodiversity of habitats, while art professors could choose trails in proximity to arts sites.

Our classifications are scalable, so one could do similar classifications for other factors once more data about MacLeish are available. Implementing this guidebook would be worth the small investment in time and money because it makes MacLeish easier to navigate, and helps visitors make better-informed trail choices according to their preferences, optimising the use of their time, and gaining a more individualised and meaningful experience. For a field station like MacLeish, being able to attract many visitors is key to its sustainability, development, and contribution to academia, which simultaneously benefits Smith, as MacLeish is an educational and extra-curricular resource for faculty and students. Furthermore, making such a guidebook wouldn’t meet any major obstacles (as we have all the necessary data), or have a detrimental environmental impact (since we are not making any physical changes).

[^github]: https://github.com/cathyleee/sds192-mp3
